# Base R

R comes with a number of functions that allow you to maniuplate, summarise and analyse data. Collectively these are known as **Base R**.

This chapter summarises some of the core functions for manipulating data in base R.

## Loading in data

The most common process to load in data is to load in a CSV file.

```{r sec4_1,echo=FALSE}
#Load in data
swfc_16_init <- read.csv("data/SWFC_2016_Machine_Readable.csv")
swfc_16 <- swfc_16_init
```

```{r sec4_2,eval=FALSE}
#Load in data
swfc_16_init <- read.csv("data/SWFC_2016_Machine_Readable.csv")
swfc_16 <- swfc_16_init
```

<div class="tip"><b>Tip:</b>
Load in and create and initial, raw version of the file, and never do anything to that object. This means if you muck anything up, you'll always have a clean dataset to start again from. This is particularly important when loading in big dataframes, such as those from SQL.
</div>
<br/>

## Basic dataframe functions

There are a number of functions which can be used to gain a summary of data: 

```{r sec4_3,eval=FALSE}
#Basic dataframe exploration functions
summary(swfc_16) #Get a summary of each variable
```

<div class="tip"><b>Tip:</b>
This worksheet is built using an add on to R called R Markdown, which integrates code, text, and images. We'll come on to how to create R Markdown documents later, but for now, the box below with a white background and grey border is an output from the above code as it renders in R Markdown. Remember this, it will appear a lot in the worksheet.
</div>
<br/>

```{r sec4_4,echo=FALSE}
#Basic dataframe exploration functions
summary(swfc_16[,c(1:6)]) #Get a summary of each variable
```

```{r sec4_5,eval=FALSE}
head(swfc_16) #Get the top 6 rows
```

```{r sec4_6,echo=FALSE}
head(swfc_16[,c(1:6)])
```

```{r sec4_7,eval=FALSE}
tail(swfc_16) #Get the bottom 6 rows
```

```{r sec4_8,echo=FALSE}
tail(swfc_16[,c(1:6)])
```

```{r sec4_9}
colnames(swfc_16) #Get list of column names
```

<div class="tip"><b>Tip:</b>

Click on the name of a dataframe to open it in a new window.
Click on the arrow to the left of it to see the structure of it, or type <span class="code">str(DATAFRAME_NAME)<span class="code">
</div>
<br/>s

We can also change column names using the above function:

```{r sec4_10}
#Identify the column number using colnames(swfc_16) and then specify the string to change it to
colnames(swfc_16)[11] <- "Region"

#...and change it back again
colnames(swfc_16)[11] <- "Government_Office_Region_Name"
```

## Selecting certain columns/removing columns

Selecting certain columns is really helpful for creating subset dataframes. Below we select the school's LA Establishment Code, its unique identifier, and all the columns to do to do with teacher absences:

```{r sec4_11}
teacher_absences <- swfc_16[,c(2,53:55)]
```

Let's break this down:

1. <span class="code">teacher_absences</span> is the name of the new dataframe we're going to create
2. We've seen the get sign before
3. <span class="code">swfc_16</span> is the dataframe we're going to select columns from
4. <span class="code">[]</span> is for selecting a certain element from an object
5. <span class="code">c()</span> stipulates a character string, in this instance a load of numbers
6. <span class="code">x:y</span> means all numbers between and including x and y
7. The comma and  nothing before it signifies that all rows must be included - the format is <span class="code">dataframe[row conditions, column conditions]</span>

Whilst we can use this method to remove columns, we can also remove individual columns using a simpler method. Say for instance we wanted to remove the 2016 LA Establishment Code, the third column:

```{r sec4_12}
swfc_16 <- swfc_16[,-3]
```

## Conditional Selections

We can select subsets of dataframes based on certain conditions. There are a number of ways to do it, but this method uses functions in the basic R set of functions, known as 'base R':

```{r sec5_1}
#Conditionally select primary schools
swfc_16_pri <- swfc_16[swfc_16$School_Phase == "Primary",]
```

Let's break this down:

1. <span class="code">swfc_16_pri</span> is the name of the new dataframe we're going to create
2. We've seen the get sign before
3. <span class="code">swfc_16</span> is the dataframe we're going to conditionally select rows from
4. <span class="code">[</span>] is for selecting a certain element from an object
5. <span class="code">$</span> is for extracting an element by name, in this instance the <span class="code">School_Phase</span> column
6. <span class="code">=="Primary"</span> signifies that rows must equal <span class="code">Primary</span>
7. The comma and then nothing after it signifies that all columns must be included - the format is <span class="code">dataframe[row conditions, column conditions]</span>

<div class="activity"><b>Activity A5.1:</b> Use conditional selections to create a new dataframe which contains all schools whose school type is an academy. Open the dataframe to see what an academy is actually called in School_Type.
</div>
<br/>

<div class="tip"><b>Tip:</b>

After typing the dollar sign when looking for dataframe column names, pause, and a dropdown list will appear.
</div>
<br/>

We can also use selections on numerical variables too:
```{r sec5_2}
swfc_16_male <- swfc_16[swfc_16$Perc_Male_Teachers > 50,]
```

However, we're not limited to just one condition:
```{r sec5_3}
#Conditionally select schools where Pupil:Teacher Ratios are below 20 and above or equal to 10
swfc_16_ptr <- swfc_16[(swfc_16$Pupil_Teacher_Ratio < 20 & swfc_16$Pupil_Teacher_Ratio >=10),]

#Conditionally select schools where Pupil:Teacher Ratios are below 10 or their LA is in Camden
swfc_16_ptr_camden <- swfc_16[(swfc_16$Pupil_Teacher_Ratio < 10 | swfc_16$LA_Name == "Camden"),]
```

<div class="activity"><b>Activity A5.2:</b> Select all schools whose StatutoryLowAge is higher than 5 and have no full time posts vacant (the fourth from last column).
</div>
<br/>

## Altering data in dataframes

Editing dataframes is a key skill. We can edit the data within columns, or create new ones.

Here we edit the <span class="code">Religious_Character</span> to be <span class="code">TRUE</span> or <span class="code">FALSE</span>. The <span class="code">Religious_Character</span> column is a column of factors - strings limited to a certain number of entries. We will first turn it into a column which can contain any string, called a character column.

<div class="tip"><b>Tip:</b>

If a column is a factor, you can see what the different entries are through <span class="code">levels(dataframe$column)</span>.
</div>
<br/>

```{r sec5_4}
#Turn Religious.Character to binary
swfc_16$Religious_Character <- as.character(swfc_16$Religious_Character)

swfc_16$Religious_Character[swfc_16$Religious_Character == 'Does not apply' |
                              swfc_16$Religious_Character == 'None' |
                              swfc_16$Religious_Character == ""] <- FALSE
swfc_16$Religious_Character[swfc_16$Religious_Character != FALSE] <- TRUE
```

This uses boolean logic (<span class="code">TRUE</span> or <span class="code">FALSE</span>). It also uses <span class="code">!=</span> which means does not equal.

We can calculate a new column too. In this instance we'll work out the percentage of teaching staff that are vacancies.

```{r sec5_5}
#Calculate percentage of posts which are vacancies
swfc_16$perc_vacancies <- swfc_16$FT_Vacant_Posts/swfc_16$Tot_Teachers_HC
```

<div class="activity"><b>Activity A5.3:</b> Turn all schools which arent an LA maintained school or a special school into 'Not LA maintained'.
</div>
<br/>

## Writing data

<div class="activity"><b>Activity A5.4:</b> The function for writing is <span class="code">write.csv()</span>. Use the help function (<span class="code">?write.csv</span>) to work out what the arguments are for this function
</div>
<br/>
