# General Coding

## Objects

Objects are any 'thing' that you create in R. They're shown in your Environment.

There are a huge number of objects in R. The following illstrates a numer of the core object types:

```{r sec3_1,eval=FALSE}
my_string <- 'DfE'
my_number <- 2017
my_boolean <- TRUE
my_vector <- c("a", "b", "c")
my_dataframe <- data.frame(var1 = c(1,2,3,4,5),
                           var2 = c("a","b","c","d","e"))
my_list <- list(my_string,my_number,my_boolean, my_vector, my_dataframe)

```

<div class="tip"><b>Tip:</b>

When naming objects it is useful to do the following:

1. Make the names meaningful but short e.g. we could call School Workforce Census Table `swfc`.

2. Stick to one convention. The standard for R is snake_case. See [here](http://adv-r.had.co.nz/Style.html) for further details on R Style Guide.


</div>
<br/>

To run that code you will first need to create a new script File > New File > R Script. 

You can then run line by line by having your cursor on that line or highlight chunks and then doing one of the following:

* Press CTRL + ENTER
* Click 'Run' in the top right hand corner of the script window

<div class="tip"><b>Tip:</b>

* Double click highlights term
* Triple click highlights line
* Quadruple click highlights entire script
</div>
<br/>

The <span class="code"><-</span> sign is called a get sign. It 'gets' the output from the right hand side and attributes it to the object name.

<div class="tip"><b>Tip:</b>
ALT + - is a shortcut to inputting the get sign.
</div>
<br/>

<div class="activity"><b>Activity A3.3:</b> 
What does <span class="code">typeof(OBJECT NAME)</span> do?
</div>
<br/>

<div class="activity"><b>Activity A3.4:</b> 
Now save your work in the <span class="code">R</span> folder!
</div>
<br/>

Writing the following code will remove a specified object:
```{r sec3_2,eval=FALSE}
rm(object_name)
```

Writing the following code will remove all objects:
```{r sec3_3,eval=FALSE}
rm(list = ls())
```

<div class="activity"><b>Activity A3.5:</b> 
Arguments are the bits of code inside brackets, and if there are multiple arguments they're separated by commas. 

Explain the what the argument inside the second <span class="code">rm</span> in the code above does.
</div>
<br/>

## Comments

<div class="activity"><b>Activity A3.6:</b> Type <span class="code">1+1</span> and run it, what comes up in the console? Put a <span class="code">#</span> in front of <span class="code">1+1</span>, what comes up in the console?</div>

Comments are really important for annotating code, so that you and others know exactly what the code does and why.

Here we add a description of what <span class="code">setwd()</span> does:

```{r sec3_4,eval=FALSE}
#Remove all objects
rm(list = ls())
```

<div class="tip"><b>Tip:</b>

* CTRL + SHIFT + C comments multiple lines at once
* CTRL + SHIFT + R creates a section, which you can jump between using the dropdown list in the bottom left corner of the script window
</div>
<br/>