# Maps

Maps in R are best plotted using <span class="code">ggplot</span> - which is good, because we already know how to use that! However, the new thing about maps is the sort of data we'll be using - as well as having data about certain variables, this data has a location attached to it too.

## Geospatial data

Geospatial data comes in three forms:

* Polygons (shapes) - Two dimensions
* Lines - one dimension
* Points - no dimensions

Polygons and points are the most common (geospatial line data only really features when relating to travel infrastructure or people movements), so we'll focus on these two.

However, before we get to plotting points or polygons we need to go through some geospatial data basics.

```{r,eval=FALSE}
cities <- "data/SHPs/england_cities.shp" %>% readOGR()

england <- "data/SHPs/England_LA_2016.shp" %>% 
  readOGR()

plot(england) +
  plot(cities,add=T,col="red")

cities <- spTransform(cities, CRS("+init=epsg:27700"))

england <- "data/SHPs/England_LA_2016.shp" %>% 
  readOGR() %>% 
  fortify(region="LA15NM") %>% 
  as.tbl() %>% 
  dplyr::select(Easting = long,
                Northing = lat,
                LAD16NM = id,
                group)

cities <- "data/SHPs/england_cities.shp" %>% 
  readOGR() %>%
  spTransform(CRS("+init=epsg:27700")) %>% 
  data.frame() %>% 
  select(City,
         Easting = coords.x1,
         Northing = coords.x2)

ggplot()+
  geom_polygon(data=england,
               aes(Easting,Northing,group=group),
               col="grey",
               fill=NA) +
  geom_point(data=cities,
             aes(Easting,Northing),
             col="red") +
  coord_equal() +
  theme_minimal() +
  geom_text(data=cities,
            aes(Easting,Northing,label=cities$City),
            check_overlap = TRUE,
            col="blue",
            hjust = 1.1,
            vjust=0.3)
```

## Point maps
```{r,eval=FALSE}
Wiltshire <- "data/SHPs/England_LA_2016.shp" %>% 
  readOGR() %>% 
  subset(LA15NM == "Wiltshire") %>% 
  fortify(region="LA15NM") %>% 
  as.tbl() %>% 
  dplyr::select(Easting = long,
                Northing = lat,
                LA15NM = id,
                group)
  
ggplot() +
  geom_polygon(data=Wiltshire,aes(Easting,Northing),col="grey",fill=NA) +
  coord_equal() +
  theme_minimal() +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
        legend.title=element_blank())

wiltshire_schools <- "data/wiltshire_schools.csv" %>% read_csv()

wiltshire_schools %>% write_csv("wiltshire_schools.csv")

ggplot() +
  geom_point(data=wiltshire_schools,aes(Easting,Northing,size=Tot_Workforce_HC,
                                        col=School_Phase)) +
  geom_polygon(data=Wiltshire,aes(Easting,Northing),col="grey",fill=NA) +
  coord_equal() +
  theme_minimal() +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
        legend.title=element_blank())
```

## Chloropleth maps

```{r,eval=FALSE}
library(rgdal)

England <- readOGR("data/SHPs/England_Regions.shp")
plot(England)

region_ave_salary <- "data/region_ave_salary.csv" %>% 
  read_csv()

Reg_England_df <- England %>% 
  fortify(region="rgn16nm") %>% 
  as.tbl() %>% 
  dplyr::select(Easting = long,
                Northing = lat,
                Region = id,
                group) %>% 
  left_join(region_ave_salary,c("Region"="Government_Office_Region_Name"))

ggplot() +
  geom_polygon(data=Reg_England_df,aes(Easting,Northing,
                                       group=group,
                                       fill=ave_sal),
               col="white") +
  coord_equal() +
  theme_minimal() +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
        legend.title=element_blank())
```
