--- 
title: "Introduction to R at DfE"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This resource is an introduction to the R programming language. It can be used as standalone learning tool, or as part of training sessions."
---

```{r setup, echo = FALSE}
library(tidyverse)
library(rgdal)
library(broom)
```

# Preface and motivation {-}

This resource is an introduction to the R programming language, which is a tool developed by statisticians across the world to do statistics, data analytics, create graphics, make dashboards and write reports. This online book can be used as standalone learning tool, or as part of training sessions.

<font size="5"><b style="color:#000000";>Chapters covered:</b></font> 

1. [Getting started with RStudio](getting-started.html)
2. [Downloading data for tutorial](data.html)
3. [General Coding](general-coding.html)
4. [Loading data](loading-and-exploring-data.html)
5. [Data manipulation with **`dplyr`**](dplyr.html)
6. [Data visualisation with **`ggplot2`**](graphs.html)
7. [Statistical operations](statistical-operations.html)
8. [Iteration](iteration.html)
9. [Functions](functions.html)
10. [Maps](maps.html)
11. [RMarkdown](r-markdown.html)

\n

<font size="5"><b style="color:#000000";>The art of the possible - show me what R is capable of</b></font>

The below links hopefully motivate some of the possible thing we can do with R, the first two are built in house at DfE. 

- [DfE Exclusion Statistics app](https://department-for-education.shinyapps.io/exclusion-statistics/) built in RShiny

- [DfE Exclusion Statistics statistical publication](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/645075/SFR35_2017_text.pdf) produced in RMarkdown

- [Blogs on making beautiful charts](https://rud.is/b/category/charts-graphs/)

- [Amazing map of Switzerland](https://timogrossenbacher.ch/2016/12/beautiful-thematic-maps-with-ggplot2-only/) build in ggplot2 

- And finally Tom's favourite, [a gif of hurricane Irma hitting Florida](https://owi.usgs.gov/vizlab/hurricane-irma/) built in ggplot2

\n

<font size="5"><b style="color:#000000";>Format of this training</b></font>

The format of the training is as follows:

* There will be a section of text describing what we will be doing

* Then there will be a code chunk showing what we will be typing, e.g.


```{r sec0_2, eval=FALSE}
print("Hello world")
```

* There'll also be activities for you to do to put your learning to the test, e.g. 

<div class="activity"><b>Activity A0.1:</b> Write a line of code that prints the string "Hello world"</div>
<br/>

* Finally, we'll also offer tips to make writing code easier

<div class="tip"><b>Tip:</b><br/> R is better than Excel</div>
