--- 
title: "DfE R training"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This resource is an introduction to the R programming language. It can be used as standalone learning tool, or as part of training sessions."
---

```{r setup, echo = FALSE}
library(tidyverse)
library(rgdal)
library(broom)
```

# Preface {-}

This resource is an introduction to the R programming language. It can be used as standalone learning tool, or as part of training sessions.

It takes you through from the first steps of opening RStudio through using it for visualisations such as graphs and maps - basically an introduction to using R for simple but effective data management, manipulation, calculations, and visualisation.

It does however work on the assumption that [R and R Studio](https://www.rstudio.com/products/rstudio/download/) are already installed.

The format of the training is as follows:

* Text describing what we will be doing

* Code showing what needs typing in:

```{r sec0_2, eval=FALSE}
print("Hello world")
```

* Activities for you to do to put your learning to the test

<div class="activity"><b>Activity A0.1:</b> Write a line of code that prints the string "Hello world"</div>
<br/>

* Little tips to make writing code easier

<div class="tip"><b>Tip:</b><br/> R is great tool</div>